Place of death in NRS COVID-19 deaths

### 2021-02-09 covid_nrs_place_of_death.Rmd
finally figured out moving averages - my approach now allows to specify how many weeks to average over, now the final test - plotting!

weekly_deaths_hb <- 
  read_excel(weekly_deaths_hb_file, sheet = 2, skip = 2) %>% 
  clean_names %>%
  rename(week_number = week_of_occurrence, ref_area = health_board, place_of_death = location_of_death, number_of_deaths = deaths) %>%
  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
  mutate(
    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
    week_number = parse_integer(str_sub(week_number, 4, 5)),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(
      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
      fill = list(number_of_deaths = 0)
      ) %>%
  # complete_weekly_deaths_with_implicit_zeros() %>%
  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
```

@@

past_deaths_hb <- 
  readxl::read_excel(past_deaths_hb_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(
    week_number = week_of_occurrence,
    place_of_death = location
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    health_board, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

past_deaths_la <-
  readxl::read_excel(past_deaths_la_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(
    week_number = week,
    place_of_death = location_of_death
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    council_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

@@

## check that the results are the same as overall deaths
map2_dfr(
  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_hb, past_deaths_sex_age),
  .y = c("overall","health_board","council_area","sex_age"),
  .f = function(data_tbl, name) {
    data_tbl %>%
      mutate(dataset=name,year=as.numeric(year)) %>%
      filter(!is.na(year) & place_of_death!="All") %>%
      group_by(dataset,year) %>%
        summarise(total = sum(number_of_deaths), .groups = "drop")
  }
) %>% 
  pivot_wider(names_from=year,values_from=total) %>%
  knitr::kable(caption = "Number of historical deaths by year as calculated from different datasets. The overall dataset and age/sex breakdown agree in figures, and the HB and LA breakdowns agree, but the two pairs differ from each other. The difference is in both directions, some years there are more deaths recorded in one pair, and some years in the other. This may be due to the following (verbatim from data description online): Health Board and Local Authority figures include non-residents. Deaths are allocated to areas based on the usual residence of the deceased. If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.")
```

@@

The average will be computed to the "left" of the current value (a week's moving average is the average of values at this week + n-1 past weeks)

# TODO: implement rolling averages also - I think this needs to be applied at source tbls first (i.e. max of sum over two weeks rather than mean of maxes?)
  # group_by(ref_area, sex, age, place_of_death) %>% 
  # mutate(across(.cols = matches("deaths_"), .fns = ~zoo::rollmean(x = .x, k = 2, align = "right", fill = NA)))

# weekly_deaths_sex_age
# weekly_deaths_hb %>%
#   group_by(year, week_number, )
```

@@

historical_average_hb <-
  past_deaths_hb %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, health_board
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

historical_average_la <-
  past_deaths_la %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, council_area
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

@@

historical_average_age <-
  past_deaths_sex_age %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by_all %>%
  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop") %>%  # recompute the sum of deaths over sexes
  group_by(
    week_number, place_of_death, age
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

historical_average_sex <-
  past_deaths_sex_age %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by_all %>%
  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & age
  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop") %>%  # recompute the sum of deaths over ages
  group_by(
    week_number, place_of_death, sex
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
```

@@

@@

merged_deaths_hb <-
  weekly_deaths_hb %>%
  inner_join(historical_average_hb, by = c("week_number","ref_area"="health_board","place_of_death"))

merged_deaths_la <-
  weekly_deaths_la %>%
  inner_join(historical_average_la, by = c("week_number","ref_area"="council_area","place_of_death"))

merged_data_sex_age <-
  weekly_deaths_sex_age %>%
  inner_join(historical_average_sex_age, by = c("place_of_death", "week_number", "sex", "age"))

merged_data_sex <-
  weekly_deaths_sex_age %>%
  filter(age!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop") %>%
  inner_join(historical_average_sex, by = c("place_of_death", "week_number", "sex"))

merged_data_age <-
  weekly_deaths_sex_age %>%
  filter(sex!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop") %>%
  inner_join(historical_average_age, by = c("place_of_death", "week_number", "age"))
```

>>>>

weekly_deaths_hb <- 
  read_excel(weekly_deaths_hb_file, sheet = 2, skip = 2) %>% 
  clean_names %>%
  rename(week_number = week_of_occurrence, ref_area = health_board, place_of_death = location_of_death, number_of_deaths = deaths) %>%
  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
  mutate(
    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
    week_number = parse_integer(str_sub(week_number, 4, 5)),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(
      nesting(year, week_number, week_number_run_over, date_w_c),  # these are 'identifying' variables that shouldn't be combinatorily exploded
      ref_area, cause_of_death, place_of_death,  # these are the variables we want to have all combinations of, and missing combinations are set to 0 deaths
      fill = list(number_of_deaths = 0)
      ) %>%
  # complete_weekly_deaths_with_implicit_zeros() %>%
  recode_place_of_death() %>%  # this needs to be done before recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()

## compute weekly data for sex and age separately
## sum over sex or age, respectively
weekly_deaths_sex <-
  weekly_deaths_sex_age %>%
  filter(age!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, sex, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")

weekly_deaths_age <-
  weekly_deaths_sex_age %>%
  filter(sex!="all") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, age, place_of_death) %>%
  summarise(across(starts_with("deaths_"), ~sum(.x)), .groups = "drop")
```

@@

past_deaths_hb <- 
  readxl::read_excel(past_deaths_hb_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(  # same format as current data
    week_number = matches("week"),
    place_of_death = matches("location"),
    ref_area = health_board
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    ref_area, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

past_deaths_la <-
  readxl::read_excel(past_deaths_la_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(  # same format as current data
    week_number = matches("week"),
    place_of_death = matches("location"),
    ref_area = council_area
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    ref_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

@@

## compute weekly data for sex and age separately
## sum over sex or age, respectively
past_deaths_sex <-
  past_deaths_sex_age %>%
  group_by_all %>%
  ungroup(age, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")

past_deaths_age <-
  past_deaths_sex_age %>%
  group_by_all %>%
  ungroup(sex, number_of_deaths) %>%  # this is done by grouping over everything except the number of deaths & sex
  summarise(number_of_deaths = sum(number_of_deaths), .groups = "drop")
```

### Do we get the same number of deaths between datasets?

```{r}
## check that the results are the same as overall deaths
map2_dfr(
  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_hb, past_deaths_sex_age),
  .y = c("overall","health_board","council_area","sex_age"),
  .f = function(data_tbl, name) {
    data_tbl %>%
      mutate(dataset=name,year=as.numeric(year)) %>%
      filter(!is.na(year) & place_of_death!="All") %>%
      group_by(dataset,year) %>%
        summarise(total = sum(number_of_deaths), .groups = "drop")
  }
) %>% 
  pivot_wider(names_from=year,values_from=total) %>%
  knitr::kable(caption = "Number of historical deaths by year as calculated from different datasets. The overall dataset and age/sex breakdown agree in figures, and the HB and LA breakdowns agree, but the two pairs differ from each other. The difference is in both directions, some years there are more deaths recorded in one pair, and some years in the other. This may be due to the following (verbatim from data description online): Health Board and Local Authority figures include non-residents. Deaths are allocated to areas based on the usual residence of the deceased. If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.")
```

@@

@@

historical_average_hb <-
  past_deaths_hb %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, ref_area
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

historical_average_la <-
  past_deaths_la %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, ref_area
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

@@

historical_average_age <-
  past_deaths_age %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, age
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

historical_average_sex <-
  past_deaths_sex %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, sex
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
```

@@

The average will be computed to the "left" of the current value (a week's moving average is the average of values at this week + n-1 past weeks, or "right" aligned).

Moving averages for historical data will be computed over the previously computed weekly means & ranges. Computing moving averages over historical ranges produces wider spread data than computing ranges of moving averages (see `order_of_min_max_rolling_mean.R`).

## compute moving average over weekly data, using average over k weeks
## using historical data to compute the moving average in the first few weeks of current data
## Note: historical data needs to be the non-summarised (raw) data!
## to compute moving average in first few week(s) of year, we add in the previous years' data and code the weeks so that they retain the right order
## i.e. in 2020 the week_number_run_over starts at 1, and the last week of 2019 is coded as 0, the 2nd last as -1, etc.
## once the moving average has been computed, we remove the past data;
## this approach works for 2021 data seamlessly, using the last weeks of 2020 for the moving average in 2021!
compute_moving_average_for_current_data <- function(current_data, historical_data) {
  bind_rows(
    historical_data %>%  # add historical data and format it so that it matches current data
      filter(year == 2019) %>%  # keep only last year before 2020
      rename(deaths_all_causes = number_of_deaths) %>%  # in historical deaths there is no split between covid/non-covid so we rename to all cause deaths
      mutate(week_number_run_over = -(52-week_number)),
    current_data
  ) %>%
    arrange(week_number_run_over) %>%  # just in case the order got messed up, re-order by week_number_run_over for the computation
    group_by_all() %>% ungroup(matches("^deaths_|week|date|year")) %>%  # clever way of getting grouping by ref_area, sex, age, whatever is available
    mutate(across(
        .cols = matches("^deaths_"), 
        .fns = ~zoo::rollmean(.x, k=weeks_to_average, fill=NA, align="right"), 
        .names = "ma{weeks_to_average}w_{.col}"
        )
    ) %>%
    ungroup %>%
    filter(year > 2019)  # keep only current data
}

rolling_deaths_hb <-
  compute_moving_average_for_current_data(current_data = weekly_deaths_hb, historical_data = past_deaths_hb)

rolling_deaths_la <-
  compute_moving_average_for_current_data(current_data = weekly_deaths_la, historical_data = past_deaths_la)

rolling_deaths_sex_age <-
  compute_moving_average_for_current_data(current_data = weekly_deaths_sex_age, historical_data = past_deaths_sex_age)

rolling_deaths_age <-
  compute_moving_average_for_current_data(current_data = weekly_deaths_age, historical_data = past_deaths_age)

rolling_deaths_sex <-
  compute_moving_average_for_current_data(current_data = weekly_deaths_sex, historical_data = past_deaths_sex)

## inspect the calculations  
rolling_deaths_la %>% filter(week_number_run_over %in% c(1:3,52:54) & str_detect(ref_area,"Glasgow") & place_of_death=="Hospital") %>% select(year, week_number_run_over, ref_area, place_of_death, deaths_all_causes, ma2w_deaths_all_causes) %>% knitr::kable(caption="Showing how rolling average was computed: week 1 is average of week 52 of 2019 (not shown) and week 1 of 2020; week 2 is average of weeks 1 & 2, etc.")

rolling_deaths_sex_age %>% filter(week_number_run_over %in% c(1:3,52:54) & str_detect(sex,"F") & age == "65-74" & place_of_death=="Hospital") %>% select(year, week_number_run_over, sex, age, place_of_death, deaths_all_causes, ma2w_deaths_all_causes) %>% knitr::kable(caption = "Showing how rolling average was computed: week 1 is average of week 52 of 2019 (not shown) and week 1 of 2020; week 2 is average of weeks 1 & 2, etc.")

## helper function that takes historical data and computes a moving average over k weeks
## with the additional trick of using the last week of all the years as the week prior to the first week
## strictly speaking we should be using the last week(s) of 2014 to compute the moving average for the first week(s) in 2015, depending on k; and we shouldn't be using the last week of 2019 in the computation either
## however, the overall effect of this is small since we're averaging over the entire 2015-2019 period
## the inaccuracy of this approach increases with increasing k
## the solution for fixing this is straightforward - just download 2014 data as well, and merge the "current" data and the historical data, and then perform the moving average calculation as usual
compute_moving_average_for_historical_data <- function(historical_data) {
  last_week_in_previous_year_required <- 53 + 2 - weeks_to_average  # this computes the last week needed from the previous year: if averaging over 2 weeks, we need week 3, if over 3 weeks, the lat week needed is 52, etc.
  bind_rows(
    historical_data %>%  # keep last week from all years and convert it to week 0
      filter(week_number %in% last_week_in_previous_year_required:53) %>% 
      mutate(week_number = week_number - 53),  # set week number for data in previous to negative numbers but retaining order 
    historical_data
  ) %>%
  arrange(week_number) %>%  # sort in ascending order of week_number
  group_by_all() %>% ungroup(matches("^deaths_"),week_number) %>%
  mutate(across(
      .cols = matches("^deaths_"), 
      .fns = ~zoo::rollmean(.x, k=weeks_to_average, fill=NA, align="right"), 
      .names = "ma{weeks_to_average}w_{.col}"
      )
  ) %>%
  ungroup %>%
  filter(week_number!=0)
}

historical_average_hb <- compute_moving_average_for_historical_data(historical_average_hb)
historical_average_la <- compute_moving_average_for_historical_data(historical_average_la)
historical_average_sex_age <- compute_moving_average_for_historical_data(historical_average_sex_age)
historical_average_sex <- compute_moving_average_for_historical_data(historical_average_sex)
historical_average_age <- compute_moving_average_for_historical_data(historical_average_age)
```

@@

merged_deaths_hb <-
  # weekly_deaths_hb %>%
  rolling_deaths_hb %>%  # use data with rolling movingincluded
  inner_join(historical_average_hb, by = c("week_number","ref_area","place_of_death"))

merged_deaths_la <-
  # weekly_deaths_la %>%
  rolling_deaths_la %>%  # use data with rolling movingincluded
  inner_join(historical_average_la, by = c("week_number","ref_area","place_of_death"))

merged_data_sex_age <-
  # weekly_deaths_sex_age %>%
  rolling_deaths_sex_age %>%  # use data with rolling movingincluded
  inner_join(historical_average_sex_age, by = c("place_of_death", "week_number", "sex", "age"))

merged_data_sex <-
  rolling_deaths_sex %>%
  inner_join(historical_average_sex, by = c("place_of_death", "week_number", "sex"))

merged_data_age <-
  rolling_deaths_age %>%
  inner_join(historical_average_age, by = c("place_of_death", "week_number", "age"))
```
